Extending Graphical Representations for Compact Closed Categories with Applications to Symbolic Quantum Computation

### 2008-05-04 papers/reasoning-graphically/paper.tex
fixed one last thing
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1482 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
final small edits.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1483 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\begin{definition}
\label{open-embedding-def}
We call a graph morphism $f$ an \emph{open embedding which is stirct for} $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
\end{definition}

\noindent The intuition behind this definition is that the subgraph of
$G$ determined by $V'$ should be preserved exactly by $f$; 
whereas other vertices may be identified and may contain additional
vertices.

@@

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_B \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Further,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down by reflection in the x-axis.  We
use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
Hadamard gate. The adjoints for $H$, $\alpha_X$ and
$\alpha_Z$ will be defined equationally.  The free compact closed
category is then given by all graphs formed by composing and tensoring
these basic graphs.

@@

We emphasise that this is a notation is for representing quantum
processes, rather than quantum states.  In this setting a state is
simply a process with no inputs; that is, a concrete graph with
empty domain.
  Since our formalism is based on the
underlying mathematical structure rather than any particular model of
quantum computation, it is capable of representing  quantum circuits,
and measurement-based quantum computations, among other models.
Indeed, an important application is to show that states or 
computations implemented differently are equivalent.

@@

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
\end{array}\qquad
\begin{array}{ccc}
\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider}\caption{ [Left] An informal equation on graphs that
  expresses the Spider Theorem.  [Right] The Spider Theorem expressed
  formally using graph patterns.  The !-boxes are named $i$ and $j$.
  The variable nodes are white and named $a$ and $b$. The non-variable
  node data (the angle) is written inside the node when non-zero.}

\end{figure}

Hence any connected subgraph involving nodes of only one colour may be
collapsed to a single vertex, with a single value $\alpha$, giving a
``spider''. Informally, this can be depicted graphically as the
equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with the sum of the $\alpha$s.
Furthermore, one can derive, from the spider theorem, $n$-fold versions of
many of the other equations.

Spiders offer a very intuitive way to manipulate graphs, and are far
more compact and convenient in calculations than the graphs built up
naively from the generators.  However, formalising spiders requires
moving from finite graphs, where each vertex has bounded degree, and
which are subject to a finite number of rewrite rules, to a system
where nodes may have arbitrarily many edges, and there are infinitely
many concrete rewrite rules. The desire to retain these intuitive
reasoning methods motivates the extension from concrete graphs to
\emph{graph patterns}. These will comprise the main subject of this
paper.

@@

\noindent However, composition with semi-circles ($d_Q$ and $e_Q$ from
Definition~\ref{compactcat-def}), on the co-domain or domain, allows
an equation involving one of the above to easily lead to a derivation
corresponding to either of the other two: proving any of the above
allows a trivial derivation of the others.

@@

It is easy to verify that $\leq_v$ is a preorder: i.e. that
matching is transitive and reflexive.

@@

More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where $G$ is a
graph and $\cal B$ is a set of subsets of $V$ such that  whenever
$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$,\footnote{One
  could consider more expressive notions of nested, or overlapping,
  node sets in the !-boxes. While such expressivity is 
  interesting,  it is not required for the system we formalise here.}  and
subject to the further restriction that no edge may connect vertices
in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
t(e)$ then $b_1 = b_2$.  
To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from its nodes, we introduce
\emph{!-box matching}.   This binary relation, written infix as
$\leq_!$, is defined such that 
$(G,{\cal B}) \leq_! (H, {\cal C})$ whenever $(H, {\cal C})$ can be
obtained from $(G,{\cal B})$ be a sequence the following operations:

@@

\begin{figure}[t]
$$\begin{array}{ccccccc}
\inlinegraphic{4em}{bbox-ex1} &  
\stackrel{\leq}{\stackrel{\ }{\small{copy}}} 
& \inlinegraphic{4em}{bbox-ex2}  & 
\stackrel{\leq}{\stackrel{\ }{\small{merge}}} 
& \inlinegraphic{4em}{bbox-ex3}  & 
\stackrel{\leq}{\stackrel{\ }{\small{drop}}} 
& \inlinegraphic{4em}{bbox-ex4}
\end{array}$$ 
\label{fig:bang-box-example}\caption{An illustration of !-box graph
  matching using the !-box operations. This involves first copying
  !-box $i_0$ twice, then merging $i_0$ and $i_1$ and finally dropping
  $i_3$.}
\end{figure}

@@

We give a formal semantics to !-box graphs by defining them in terms
of a set of graphs in the underlying representation. We denote the
interpretation of a !-box graph $(G,{\cal B}$ by $\binterp{(G,{\cal
    B})}$ and say that members of $\binterp{G}$ are instances of $G$.

\begin{definition}
  The \emph{interpretation} $\binterp{(G,{\cal B})}$ is a
  set of concrete graphs defined 
  \[
  \binterp{(G,{\cal B})} = \{ H \;\;|\;\; (G,{\cal B}) \leq_! (H,\emptyset) \}
  \]
  i.e. the those graphs matched by the !-box graph that have no
  !-boxes.
\end{definition}

Observe that every concrete instance of a !-box graph can be defined
by pairing each !-box with the natural number that defines how many
copies are made of it. Thus $\binterp{G}$ is isomorphic to the set of
$k$-tuples of natural numbers, where $k$ is the number of !-boxes.

@@

The need for the !-box matching operation, rather than using a direct
$k$-tuple interpretation, is to allow matching between !-box graphs,
and thus to provide a mechanism for derived rules.

@@

The representation of Compact Closed Categories as graphs, discussed
in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, graphical rules such the spider
theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
not expressible. In this section we combine the {\em !-boxes} and {\em
  variable-node} extensions to graphs. We call this representation
\emph{graph patterns}. This forms a representation that allows us to
express, in a finite way, the infinite family of concrete graphs
equations. In particular, the Spider Theorem can now be represented as
shown in Figure~\ref{fig:spider}-right. We also extend the notion of
matching for graph pattens.
%as well as present a corresponding notion of plugging. 
This provides the foundations for the rewriting machinery
in~\S\ref{sec:rewriting} which can then be used to reasoning about
quantum computation.

@@

In this section we describe how the graph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. An object-level formalisation defines a set of rules which
are treated as the axioms for the system; for instance, the equation
from Figure~\ref{}. It also defines the data at the nodes and edges as
well as corresponding data-matching behaviour.  For its part, the
meta-level provides generic machinery to manipulate graphs and derive
new rules.  We now describe the meta-level framework, noting the
conditions for a rule to be valid, and prove the systems adequacy for
rewriting. The resulting system forms the basis for an interactive
proof assistant that supports reasoning about compact closed
categories.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of graph patterns. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote a set of concrete equational rules.

@@

\item There has to be a isomorphism between variable nodes in the left
  and right hand subgraphs. Given a matching on the left hand side,
  the target subgraph is replaced with the right hand side while
  keeping the same instantiations for the isomorphic variable nodes.

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{fig:spider} shows the Spider Theorem. In this
figure, the mapping between !-boxes is represented by !-boxes having
the same name on the left and right of a rule.  Similarly, the
isomorphism between variable nodes is captured by the set of variable
node names on the left and right hand side being equal.

@@

\noindent $\Gamma$ is the set of object-level axioms.

We assume that axioms in $\Gamma$ meet the validity conditions
described earlier. These rules all preserve the validity conditions on
equations and thus the system as a whole ensures only valid rules are
derived.

The reflexivity rule (\emph{refl}) assumes that $A$ is a well-formed
pattern graph. This rule allows a new graph to be introduced. By then
applying the \emph{subst} rule, intermediate results are derived which
can themselves be used to rewrite other rules and conjectures.  In
this way, the system allows derived rules to provide an abbreviation
for a combination of steps.

A sets of rules can be applied automatically to simplify a graph or
simulate computation in an the object domain. For this to terminate, a
suitable left-to-right ordering on rules can be observed such as the
size of the subgraph must decrease. In \S\ref{sec:case-study} we
illustrate simulating a quantum computation.

@@

The model of quantum computation introduced in
\S\ref{sec:quotients-rewriting} provides an object formalism for our
meta-level framework. In particular, the object level axioms come from
the lifting the equations in Figure~\ref{fig:graph-quant-eqns} and the
from the encoding of the spider law as shown in
Figure~\ref{fig:spider}.

@@

To allow composition of rules to compute the resulting angles we give
the $X$ and $Z$ nodes an \emph{angle expression}. When a node is
within a !-box, the expression is a single \emph{angle-variable} which
gets instantiated to a new angle-variable in each of the unfoldings of
the !-box.  When a node is not within a !-box, the angle-expression is
a mapping from a set of angle-variables to the corresponding rational
coefficient. When an angle-expression contains a angle-variable within
a !-box, this is interpreted as a sum of the variables that result
from its unfolding.  This rather simple expression language has a
normal form by ordering the angle-variable by name.  Matching then
results in angle-variables being instantiated and the expressions in
all affected nodes are then re-normalised. An additional
implementation detail must also be observed for the substitution rule:
it must ensure that angle-variables in the rule being applied are
distinct from those in the expression being rewritten.

@@

The quantum Fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing a
polynomial time algorithm for factoring. In our graph pattern calculus
this circuit becomes the top-left graph in
Figure~\ref{fig:quantum-transform}. This figure shows how computation
can be symbolically performed by rewriting with the lifted equations
from Figure~\ref{fig:graph-quant-eqns} and the graph pattern version
of the Spider Theorem.

@@

We have introduced a representation for graphs which can formally
characterise the ellipses notation used informally to represent
certain infinite families of graph rewrites, such as the Spider
Theorem.  This representation provides the foundation for a simple
meta-logic for reasoning about models of compact closed categories. We
illustrated this by providing an account of quantum computation and
showing how computation can be performed. Having developed the basic
representational machinery and shown matching to be decidable, we are
left with several exciting avenues for further research.  The most
immediate direction we are pursuing is to provide a full
implementation - only a partial one is currently
available~\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
Other areas of further work include considering confluence results for
sets of rewrite rules, increasing the expressiveness of the
representation for graph-patterns, and finding a complete set of
rewrite rules for the considered model of quantum computation.

>>>>

\begin{definition}
\label{open-embedding-def}
We call a graph morphism $f$ an \emph{open embedding which is strict for} $V'$ if:
\begin{enumerate}
\item $f_E$ is injective;
\item $f_V$ restricted to $V'$ is injective; and,
\item $f$ is strict for $V'$.
\end{enumerate}
\end{definition}

\noindent The intuition behind this definition is that the subgraph of
$G$ determined by $V'$ should be preserved exactly by $f$; 
whereas other vertices may be identified and may contain additional
incident edges.

@@

Every arrow $f:A\to B$ in a compact closed category \catC
has a \emph{name} and \emph{coname}:
\[
\name{f} : I \to A^* \otimes B, \qquad \coname{f} : A \otimes  B^* \to I,
\]
which are constructed as $\name{f} = (\id{A^*}\otimes f) \circ d_A$ and
$\coname{f} = e_B \circ (f \otimes \id{B^*})$.  Hence there are natural
isomorphisms $\catC(A,B) \iso \catC(I,A^*\otimes B) \iso
\catC(A\otimes B^*,I)$ making \catC monoidally closed\footnote{In
  general compact closed categories  are models of multiplicative
  linear logic where $A \multimap B$ is defined as $A^\bot \otimes B$.}.
Furthermore,  $f$ has a dual, $f^* : B^* \to A^*$, defined by 
\[
f^* = (\id{A^*} \otimes e_B) \circ (\id{A^*}\otimes f \otimes
\id{B^*}) \circ (d_A \otimes \id{B^*})
\]
By virtue of equations \eqref{eq:comcl1} and \eqref{eq:comcl2}, $f^{**} =
f$.  Thus $(\cdot)^*$ lifts to an involutive functor
$\catC^{\text{op}} \to \catC$,  making $\catC$ equivalent to its
opposite.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] a single object, written $Q$.
\item[Arrows:] there are families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down by reflection in the x-axis.  We
use colours (light green and a darker red) to denote the two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H} and represents a
Hadamard gate. The adjoints for $H$, $\alpha_X$ and
$\alpha_Z$ will be defined equationally.  The free compact closed
category is then given by all graphs formed by composing and tensoring
these basic graphs.

@@

We emphasise that this is a notation for representing quantum
processes, rather than quantum states.  In this setting a state is
simply a process with no inputs; that is, a concrete graph with
empty domain.
  Since our formalism is based on the
underlying mathematical structure rather than any particular model of
quantum computation, it is capable of representing  quantum circuits,
and measurement-based quantum computations, among other models.
Indeed, an important application of this work is to show that states or 
computations implemented differently are equivalent.

@@

\noindent Hence any connected subgraph involving nodes of only one colour may be
collapsed to a single vertex, with a single value $\alpha$, giving a
``spider''. Informally, this can be depicted graphically as the
equation in Figure~\ref{fig:spider}-left. Conversely, a spider may be
arbitrarily divided into sub-spiders, provided the total in- and
out-degree is preserved, along with the sum of the $\alpha$s.
Furthermore, one can derive, from the Spider Theorem, $n$-fold versions of
many of the other equations.

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
\end{array}\qquad
\begin{array}{ccc}
\inlinegraphic{12em}{spider_lhs_patt} & \;=\; & \inlinegraphic{9em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider}\caption{ [Left] An informal equation on graphs that
  expresses the Spider Theorem.  [Right] The Spider Theorem expressed
  formally using graph patterns.  The !-boxes are named $i$ and $j$.
  The variable nodes are white and named $a$ and $b$. The non-variable
  node data (the angle) is written inside the node when non-zero.}
\end{figure}

Spiders offer a very intuitive way to manipulate graphs, and are far
more compact and convenient in calculations than the graphs built up
naively from the generators.  However, formalising spiders requires
moving from finite graphs, where each vertex has bounded degree, and
which are subject to a finite number of rewrite rules, to a system
where nodes may have arbitrarily many edges, and there are infinitely
many concrete rewrite rules. The desire to retain intuitive reasoning
methods for these infinite families of rewrites motivates the
extension from concrete graphs to \emph{graph patterns}, the main
subject of this paper.

@@

\noindent However, composition with semi-circles ($d_Q$ and $e_Q$ from
Definition~\ref{compactcat-def}), on the co-domain or domain, allows
an equation involving one of the above to easily lead to a derivation
corresponding to either of the other two: given any one of the allows
a trivial derivation of the others.

@@

%It is easy to verify that $\leq_v$ is a preorder: i.e. that
%matching is transitive and reflexive.

@@

More formally, a \emph{!-box graph} is a pair $(G, {\cal B})$ where
$G$ is a graph and $\cal B$ is a set of disjoint subsets of $V$ i.e.
$b_1, b_2 \in {\cal B}$ then $b_1 \cap b_2 =\emptyset$.\footnote{One
  could consider more expressive notions of nested, or overlapping,
  node sets in the !-boxes. While such expressivity is interesting, it
  is not required for the system we formalise here.}
%% -- we only need the unconnected !-box restriction for the merge operation. 
%  and
%subject to the further restriction that no edge may connect vertices
%in distinct !-boxes, i.e. if $v_1 \in b_1$, $v_2 \in b_2$ and $s(e) =
%t(e)$ then $b_1 = b_2$.  
To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from its nodes, we introduce
\emph{!-box matching}.   This binary relation, written infix as
$\leq_!$, is defined such that 
$(G,{\cal B}) \leq_! (H, {\cal C})$ whenever $(H, {\cal C})$ can be
obtained from $(G,{\cal B})$ by a sequence the following operations:

@@

\begin{figure}[t]
$$\begin{array}{ccccccc}
\inlinegraphic{4em}{bbox-ex1} &  
\stackrel{\leq}{\stackrel{\ }{\mbox{\small{copy}}}} 
& \inlinegraphic{4em}{bbox-ex2}  & 
\stackrel{\leq}{\stackrel{\ }{\mbox{\small{merge}}}} 
& \inlinegraphic{4em}{bbox-ex3}  & 
\stackrel{\leq}{\stackrel{\ }{\mbox{\small{drop}}}} 
& \inlinegraphic{4em}{bbox-ex4}
\end{array}$$ 
\label{fig:bang-box-example}\caption{An illustration of !-box graph
  matching using the !-box operations. This involves first copying
  !-box $i_0$ twice, then merging $i_0$ and $i_1$ and finally dropping
  $i_3$.}
\end{figure}

@@

We give a formal semantics to !-box graphs by defining them in terms
of a set of graphs in the underlying representation. In particular, we
denote the interpretation of a !-box graph $(G,{\cal B})$ by
$\binterp{(G,{\cal B})}$ and say that its members are instances.

\begin{definition}
  The \emph{interpretation} $\binterp{(G,{\cal B})}$ is a
  set of concrete graphs defined by
  \[
  \binterp{(G,{\cal B})} = \{ H \;\;|\;\; (G,{\cal B}) \leq_! (H,\emptyset) \}
  \]
  i.e. the those graphs matched by the !-box graph that have no
  !-boxes.
\end{definition}

Observe that every concrete instance of a !-box graph can be defined
by pairing each !-box with the natural number that defines how many
copies are made of it. Thus $\binterp{G}$ is isomorphic to the set of
$k$-tuples of natural numbers, where $k$ is the number of !-boxes.
The need for the !-box matching operation, rather than using a direct
$k$-tuple interpretation, is to allow matching between !-box graphs,
and thus to provide a mechanism for derived rules.

@@

@@

The representation of Compact Closed Categories as graphs, discussed
in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, graphical rules such the Spider
Theorem, from Figure~\ref{fig:spider}-left, are frequently needed, but
not expressible. In this section we combine the {\em !-boxes} and {\em
  variable-node} extensions to graphs. We call this representation
\emph{graph patterns}. This forms a representation that allows us to
express, in a finite way, certain infinite families of equations
between concrete graphs. In particular, the Spider Theorem can now be
represented as shown in Figure~\ref{fig:spider}-right. We also extend
the notion of matching for graph pattens.
%as well as present a corresponding notion of plugging. 
This provides the foundations for the rewriting machinery
in~\S\ref{sec:rewriting} which can then be used to reason about
quantum computation.

@@

In this section we describe how the graph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. An object-level formalisation defines a set of rules which
are treated as the axioms for the system; for instance, the equations
from Figure~\ref{fig:graph-quant-eqns}. It also defines the data at
the nodes and edges as well as corresponding data-matching behaviour.
For its part, the meta-level provides generic machinery to manipulate
graphs and derive new rules.  We now describe the meta-level
framework, noting the conditions for a rule to be valid, and prove the
systems adequacy for rewriting. The resulting system forms the basis
for an interactive proof assistant that supports reasoning about
compact closed categories.

@@

In our framework, the axioms defined by an object-level model, as well
as derived rules, are pairs of graph patterns. Such a pair represents
the left and right hand sides of an equation. Rules are declarative in
that they denote a set of concrete equational rules.

@@

\item There has to be a isomorphism between variable nodes in the left
  and right hand subgraphs. Given a matching against the left hand
  side of a rule, the target subgraph is replaced with the right hand
  side while keeping the same instantiations for the isomorphic
  variable nodes of the right hand side.

@@

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graph with unique names. For example,
Figure~\ref{fig:spider} shows the Spider Theorem. In this figure, the
mapping between !-boxes is represented by !-boxes having the same name
on the left and right of a rule.  Similarly, the isomorphism between
variable nodes is captured by the set of variable node names on the
left and right hand side being equal.

@@

\noindent where $\Gamma$ is the set of object-level axioms.

We assume that the axioms in $\Gamma$ meet the validity conditions
described earlier. These rules all preserve the validity conditions on
equations and thus the system as a whole ensures only valid rules are
derived. For the reflexivity rule (\emph{refl}), we assumes that $A$
is a well-formed pattern graph. This rule allows a new graph to be
introduced. By then applying the \emph{subst} rule, intermediate
results are derived which can themselves be used to rewrite other
rules and conjectures.  In this way, the system allows derived rules
to provide an abbreviation for a combination of steps.

A sets of rules can be applied automatically to simplify a graph or
simulate computation in the object domain. For such rewriting to
terminate, a suitable left-to-right ordering on rules needs to be
observed, such as a decrease in the size of the subgraph. In
\S\ref{sec:case-study} we illustrate simulating a quantum computation.

@@

The model of quantum computation introduced in
\S\ref{sec:quotients-rewriting} provides an object formalism for our
meta-level framework. In particular, the object level axioms come from
lifting the equations in Figure~\ref{fig:graph-quant-eqns} and from
the encoding of the Spider Theorem as shown in
Figure~\ref{fig:spider}.

@@

To allow composition of rules to compute the resulting angles we give
the $X$ and $Z$ nodes an \emph{angle expression}. When a node is
within a !-box, the expression is a single \emph{angle-variable} which
gets instantiated to a new angle-variable in each of the unfoldings of
the !-box.  When a node is not within a !-box, the angle-expression is
a mapping from a set of angle-variables to the corresponding rational
coefficient. When an angle-expression contains an angle-variable within
a !-box, this is interpreted as a sum of the variables that result
from its unfolding.  This rather simple expression language has a
normal form by ordering the angle-variable by name.  Matching then
results in angle-variables being instantiated and the expressions in
all affected nodes are then (re)normalised. An additional
implementation detail must also be observed for the substitution rule:
it must ensure that angle-variables in the rule being applied are
distinct from those in the expression being rewritten.

@@

The quantum Fourier transform is among the most important quantum
algorithms, forming an essential part of Shor's
algorithm~\cite{Shor:PolyTimeFact:1997}, famous for providing
polynomial factoring. In our graph pattern calculus this circuit
becomes the top-left graph in Figure~\ref{fig:quantum-transform}. This
figure shows how computation can be symbolically performed by
rewriting with the lifted equations from
Figure~\ref{fig:graph-quant-eqns} and the graph pattern version of the
Spider Theorem.

@@

We have introduced a representation for graphs which can formally
characterise the ellipses notation used informally to represent
certain infinite families of graph rewrites, such as the Spider
Theorem.  This representation provides the foundation for a simple
meta-logic for reasoning about models of compact closed categories. We
illustrated this by providing an account of quantum computation and
showing how computation can be performed. Having developed the basic
representational machinery and shown matching to be decidable, we are
left with several exciting avenues for further research.  The most
immediate direction we are pursuing is to provide a full
implementation - only a partial one is currently
available\footnote{\url{http://dream.inf.ed.ac.uk/projects/quantomatic}}.
Other areas of further work include considering confluence results for
sets of rewrite rules, increasing the expressiveness of the
representation for graph-patterns, and finding a complete set of
rewrite rules for the considered model of quantum computation.
